---
title: "DataCarpentry_Day2_SubsettingData"
author: "Kacie Ring"
date: "9/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("ggplot")
install.packages("plyr")
install.packages("gapminder")

```

##INTRODUCTION

R has many powerful subset operators. Mastering them will allow you to easily perform complex operations on any kind of dataset.

There are six different ways we can subset any kind of object, and three different subsetting operators for the different data structures.

##HELPFUL TERMS

#Atomic vector: 
simplest form of a vector containing either character stings (e.g. a,b,c), numbers, and logical values (TRUE/FALSE)

#Index/Indeces: 
Posotion in a vector (1, 2, 3, 4). We will be asking R to pull, remove, add the nth index




```{r}
#Let’s start with the workhorse of R: a simple numeric vector.
#x is our numeric vextor 

x <- c(5.4, 6.2, 7.1, 4.8, 7.5)

#Let's name em! 
names(x) <- c('a', 'b', 'c', 'd', 'e')
#does it work? 
x

```

#Atomic vectors 

Simple vectors containing either character stings (e.g. a,b,c), numbers, and logical values (TRUE/FALSE)

Atomic vectors cannot be further simplified 

##ACCESSING ELEMENTS USING THEIR INDICES 

```{r}
#Remeber our atominc vector? Let's extract elements of it by giving their corresponding index

#position one
x[1]

#Cool! how about position 4? 


```

It may look different, but the square brackets operator is a function. For vectors (and matrices), it means “get me the nth element”.


```{r}
#We can also ask R for multiple elements at one time with x[c(#,#)]

#Asking R to pull the first and third element in our atominc vector

x[c(1,3)]

#right on, now lets take slices the vector

x[1:4]

#the : operator creates a sequence of numbers from the left element to the right.

#similarly to 

c(1,2,3,4)


```


```{r}
#want multiple elements at one time? no problem!

x[c(1,1,3)]


```

```{r}
#what if we ask for an index beyond the length of our vector?? 

x[6]

#0th element??

x[0]

#an emtpy vecotr! 
```

##Fun fact! 
In many programming languages (C and Python, for example), the first element of a vector has an index of 0. In R, the first element is 1. Isn't the inconsistency fabulous? 

##SKIPPING AND REMOVING ELEMENTS 

```{r}
#If we use a negative number as the index of a vector, R will return every element except for the one specified:

x[-2]

#remove an index of your choice! 


#We can also skip/remive multiple elements (two different ways)

#option 1
x[c(-1,-5)]
#option 2
x[-c(1,5)]

```

```{r}
#Let's try and skip slices! 

x[-1:3]

#Whoopsie! a cryptic error message.. this is a common mistake! 
#Order of operations are very important in R 

# : is really a function. It takes its first argument as -1, and its second as 3, so generates the sequence of numbers: c(-1, 0, 1, 2, 3). We do not have -1 and 0 in our atomic vector! 

##How do we fix this? 

#The correct solution is to wrap that function call in brackets, so that the - operator applies to the result:

x[-(1:3)]


```

```{r}
#To permanetaly remove elements from a vector, we need to assign the result back into the variable:
#first let's look at out vector
x
#let's remove the 4th index
x<-x[-4]
x

```

##CHALLENGE 1

Given the following code 

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)

```

Come up with at least 2 different commands that will produce the following output:

 b   c   d 
6.2 7.1 4.8 

```{r}
#Hint slicing (:) and removing (-) will be helpful operators! 


```

You are crushing it! Let's move on 

##Subsetting by name

```{r}

#We can extract elements by using their name, instead of extracting by index:

x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5) # we can name a vector 'on the fly'
x[c("a", "c")]



```

#Why would I extract elements by name? More code for the same thing??

his is usually a much more reliable way to subset objects: the position of various elements can often change when chaining together subsetting operations, but the names will always remain the same! ;) 


##Subsetting through other logical operations 

Remember logical operators is just a fancy term for TRUE/FALSE

```{r}

#Let's use ~~~logical operators~~ to subset the 4th and 6th element

x[c(FALSE, FALSE, TRUE, FALSE, TRUE)]

#Since COMPARISON operators (e.g. >, <, ==) evaluate to logical vectors, we can also use them to succinctly subset vectors: the following statement gives the same result as the previous one.

x[x > 7]

#Breaking it down, this statement first evaluates x>7, generating a logical vector c(FALSE, FALSE, TRUE, FALSE, TRUE), and then selects the elements of x corresponding to the TRUE values.

#This is happening in the invisible programming land that we do not see 

##Your turn!! use a logical operation to subset the 1st and third element, name your NEW vector z 

```


```{r}
#want another way to do the same thing? remember our ol' reliable names! 

#We can use == to mimic the previous method of indexing by name (remember you have to use == rather than = for comparisons):

x[names(x) == "a"]



```


##TIP TIME! 

#Combining logical conditions

We often want to combine multiple logical criteria. For example, we might want to find all the countries that are located in Asia or Europe and have life expectancies within a certain range. Several operations for combining logical vectors exist in R:

## & 
The “logical AND” operator: returns TRUE if both the left and right are TRUE.

## | 

the “logical OR” operator: returns TRUE, if either the left or right (or both) are TRUE.

## ! 

the “logical NOT” operator: converts TRUE to FALSE and FALSE to TRUE. It can negate a single logical condition (eg !TRUE becomes FALSE), or a whole vector of conditions(eg !c(TRUE, FALSE) becomes c(FALSE, TRUE)).

##CHALLENGE 2

```{r}

#Given the following code: 
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)

#Write a subsetting command to return the values in x that are greater than 4 and less than 7.

#name your new subset x_subset. HINT: use the & logical condition 

```


##Non unique names 

You should be aware that it is possible for multiple elements in a vector to have the same name. (For a data frame, columns can have the same name — although R tries to avoid this — but row names must be unique.) Consider these examples:

```{r}
#create a vector 
x <- 1:3
x

#assign non unique names 'a'
names(x) <- c('a', 'a', 'a')
x

#what happens when you call index a? 
x['a']

#only returns the first value! 

#how do we return all three values if the have the same name? 
#like dis! 
x[names(x) == 'a']

```

##Skipped name elements 

Skipping or removing named elements is a little harder. If we try to skip one named element by negating the string, R complains (slightly obscurely) that it doesn’t know how to take the negative of a string:

```{r}

x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5) # we start again by naming a vector 'on the fly'
#let's try and get rid of "a" 

x[-"a"]

#Whoopsie! Error messgae! what the freak is a urnary operator?? 

#We can use the != (not-equals.. remeber the logical condition above?) operator to construct a logical vector that will do what we want:

x[names(x) != "a"]


```

Skipping multiple named indices is a little bit harder still. 

```{r}
#Suppose we want to drop the "a" and "c" elements, so we try this:

x[names(x)!=c("a","c")]



```

R did something, but it gave us a warning that we ought to pay attention to - and it apparently gave us the wrong answer (the "c" element is still included in the vector)!

So what does != actually do in this case? That’s an excellent question.

##Recycling 

Let’s take a look at the comparison component of this code:

```{r}


names(x) != c("a", "c")


```

Why does R give TRUE as the third element of this vector, when names(x)[3] != "c" is obviously false? 

When you use !=, R tries to compare each element of the left argument with the corresponding element of its right argument. What happens when you compare vectors of different lengths?

#SWITCH TO VISUAL 

#TBH I am a little confused by the recycling, let's learn it together :) 

#HOW DO WE FIX? 

The way to get R to do what we really want (match each element of the left argument with all of the elements of the right argument) it to use the %in% operator. The %in% operator goes through each element of its left argument, in this case the names of x, and asks, “Does this element occur in the second argument?”. Here, since we want to exclude values, we also need a ! operator to change “in” to “not in”:

```{r}
#Lets uses the %in% operator 

x[! names(x) %in% c("a","c") ]

#woo hoo! 

```

##CHALLENGE 3 

Selecting elements of a vector that match any of a list of components is a very common data analysis task. For example, the gapminder data set contains country and continent variables, but no information between these two scales. Suppose we want to pull out information from southeast Asia: 

#how do we set up an operation to produce a logical vector that is TRUE for all of the countries in southeast Asia and FALSE otherwise?

```{r}

#does everyone have gapminder data in ther enviro?? 

seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
## read in the gapminder data that we downloaded in episode 2

gapminder <- read.csv("data/gapminder_data.txt", header=TRUE)
## extract the `country` column from a data frame (we'll see this later);
## convert from a factor to a character;
## and get just the non-repeated elements

countries <- unique(as.character(gapminder$country))

```


There’s a wrong way.... (using only ==), which will give you a warning; a clunky way (using the logical operators == and |); and an elegant way (using %in%). See whether you can come up with all three and explain how they (don’t) work.

```{r}
#reminder!

#how do we set up an operation to produce a logical vector that is TRUE for all of the countries in southeast Asia and FALSE otherwise?


#Hint either use == and | or more elegantly %in%

#HINT to get you started (countries == "" | ...)

#bonus points for using %in%


```

Weird flex but ok, lets continue 

##Handling special values

At some point you will encounter functions in R that cannot handle missing, infinite, or undefined data.

#There are a number of special functions you can use to filter out this data:

#is.na 
will return all positions in a vector, matrix, or data.frame containing NA (or NaN)

#is.nan, and is.infinite 
will do the same for NaN and Inf.

#is.finite 
will return all positions in a vector, matrix, or data.frame that do not contain NA, NaN or Inf.

#na.omit 
will filter out all missing values from a vector


##Factor subsetting 

Now that we’ve explored the different ways to subset vectors, how do we subset the other data structures?

Factor subsetting works the same way as vector subsetting.

```{r}
#Let's name our factor f 

f <- factor(c("a", "a", "b", "c", "c", "d"))

#subset factor "a" 

f[f == "a"]

#let's use %in% to pull b and c 

f[f %in% c("b", "c")]

#Let's pull the first three elements in our factor 
f[1:3]

#Skipping elements will not remove the level even if no more of that category exists in the factor:

f[-3]
```


##Matrix Subsetting

Matrices are also subsetted using the [ function. In this case it takes two arguments: the first applying to the rows, the second to its columns:

#Matix defintion

A collection of elements of the same data type (numeric, character, or logical) arranged into a fixed number of rows and columns. Since you are only working with rows and columns, a matrix is called two-dimensional.

```{r}

set.seed(1) #note set.seed is a pseudorandom number generator
#name our matrix m
m <- matrix(rnorm(6*4), ncol=4, nrow=6) #4 columns & 6 rows with random #'s 
#let's look at m
m

#let's subset rows 3 through 4 and coulmns 3 and 1 
m[3:4, c(3,1)]

#You can leave the first or second arguments blank to retrieve all the rows or columns respectively:
#aka subset all rows with columns 3 & 4 
m[, c(3,4)]

#If we only access one row or column, R will automatically convert the result to a vector:
#subset row 3! 
m[3,]

#REMEMBER ROWS FIRST COLUMNS SECOND :) 

#Try it out. Go ahead and subset crows 2,4 and 5 and columns 1 through three 


```

If you want to keep the output as a matrix, you need to specify a third argument; drop = FALSE:

```{r}
#This is saying pull row 3 and keep the output as matrix 

m[3, , drop=FALSE]

#Unlike vectors, if we try to access a row or column outside of the matrix, R will throw an error:

m[, c(3,6)]

#out of bounds
#Guy Fieri 

```

Because matrices are vectors, we can also subset using only one argument:

```{r}

m[5] 
#What does m[5] correspond to?? 

#This usually isn’t useful, and often confusing to read. However it is useful to note that matrices are laid out in column-major format by default. That is the elements of the vector are arranged column-wise:

#Let's see where m[5] comes from

matrix(1:6, nrow=2, ncol=3)


```


If you wish to populate the matrix by row, use byrow=TRUE:

```{r}
#it R's default makes your brain go haywire, use the argument byrow=TRUE

matrix(1:6, nrow=2, ncol=3, byrow=TRUE)

#That looks better in my brain #rows4lyfe


```

NOTE: Matrices can also be subsetted using their rownames and column names instead of their row and column indices. You just gotta name em! 

##CHALLENGE 5 

Given the following code:
```{r}
#create a matrix with numbers 1 through 18 with 3 rows and 6 columns 

m <- matrix(1:18, nrow=3, ncol=6)
print(m)


```

Which of the following commands will extract the values 11 and 14?

A. m[2,4,2,5]

B. m[2:5]

C. m[4:5,2]

D. m[2,c(4,5)]

#TRY each of these commands below! What is the correct answer?? 

```{r}

```

Y'all are kickin ass! Let's keep going!!! 

##LIST SUBSETTING 

Now we’ll introduce some new subsetting operators. There are three functions used to subset lists. We’ve already seen these when learning about atomic vectors and matrices: [, [[, and $.

Using [ will always return a list. If you want to subset a list, but not extract an element, then you will likely use [.

```{r}
#Lets make and name a list 
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(mtcars))


xlist[1] #this returns a list with one element.


```

We can subset elements of a list exactly the same way as atomic vectors using [. Comparison operations however won’t work as they’re not recursive, they will try to condition on the data structures in each element of the list, not the individual elements within those data structures.


```{r}
xlist[1:2]
```

To extract individual elements of a list, you need to use the double-square bracket function: [[.

```{r}

xlist[[1]]

```

Notice that now the result is a vector, not a list.

You can’t extract more than one element at once:

```{r}

xlist[[1:2]]

```


Nor use it to skip elements:

```{r}

xlist[[-1]]

```


But you can use names to both subset and extract elements:

```{r}
xlist[["a"]]

```

The $ function is a shorthand way for extracting elements by name:

```{r}
xlist$data
```

##CHALLENGE 5

Given the following list:
```{r}

xlist <- list(a = "Software Carpentry", b = 1:10, data = head(mtcars))

#Using your knowledge of both list and vector subsetting, extract the number 2 from xlist. Hint: the number 2 is contained within the “b” item in the list.



```



##DATA FRAMES 

Remember the data frames are lists underneath the hood, so similar rules apply. However they are also two dimensional objects:

[ with one argument will act the same way as for lists, where each list element corresponds to a column. The resulting object will be a data frame:

```{r}
head(gapminder[3])
```

```{r}
#Similarly, [[ will act to extract a single column:

head(gapminder[["lifeExp"]])

#And $ provides a convenient shorthand to extract columns by name:

head(gapminder$year)

#With two arguments, [ behaves the same way as for matrices:

gapminder[1:3,]

#If we subset a single row, the result will be a data frame (because the elements are mixed types):
gapminder[3,]

#But for a single column the result will be a vector (this can be changed with the third argument, drop = FALSE).
```

##CHALLENGE 7 

Fix each of the following common data frame subsetting errors:

Extract observations collected for the year 1957

```{r}
Fix each of the following common data frame subsetting errors:

Extract observations collected for the year 1957

gapminder[gapminder$year = 1957,]
#Extract all columns except 1 through to 4

gapminder[,-1:4]
#Extract the rows where the life expectancy is longer the 80 years

gapminder[gapminder$lifeExp > 80]
#Extract the first row, and the fourth and fifth columns (continent and lifeExp).

gapminder[1, 4, 5]
#Advanced: extract rows that contain information for the years 2002 and 2007

gapminder[gapminder$year == 2002 | 2007,]

```


##CHALLENGE 8 

```{r}
#Why does gapminder[1:20] return an error? How does it differ from gapminder[1:20, ]?

gapminder[1:20]
  
#Create a new data.frame called gapminder_small that only contains rows 1 through 9 and 19 through 23. You can do this in one or two steps.


```

