---
title: "Data Structures"
author: "sbsambado"
date: "9/20/2021"
output: html_document
---

## Data Structures - Software Carpentry Workshop


### Agenda 

+ Big picture
+ Set up
+ Tools & Exercises 
+ Questions
+ Debrief  

**note: feel free to ask questions at any point in the lecture**

>[link to R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/04-data-structures-part1/index.html)



#### Big Picture

+ there is so much data in this world! However, *only some of it is useful* for your question
+ to make data useful for you, *you will have to manipulate it* 
+ before you can manipulate data, you must *know what you data are*


#### (some) types of data 

1. **Character** --> characteristics, like words.

  + continuous: can take on *any* value
  + discrete: can be only *certain* values

2. **Numerical** --> numbers, duh.

3. **Logical** --> TRUE/FALSE



#### rules of data

+ Each **variable** has its own **column**

+ Each **observation** has its own **row**

+ Each **value** has its own **cell**


*note*: I tend to use a certain type of language while live coding 

+ function(argument) 

+ HINT: DATASET, COLUMN, VARIABLE, OBSERVATION




#### Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


## necessary packages

# tidyverse
  



## necessary data 


  # Dataframe 1 -- cats  (we'll make it)

cats <- data.frame(coat = c("calico", "black", "tabby"), # variable 1
                    weight = c(2.1, 5.0, 3.2), # variable 2
                    likes_string = c(1, 0, 1)) # variable 3


# let's write this into a .csv

write.csv(x = cats, # data
          file = "feline-data.csv",  # file path
          row.names = FALSE) # give row number



# Dataframe 2 -- gapminder (from Carpentries)

gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", stringsAsFactors = TRUE) 

```

**Check in: Off the top of your head, what type of data are in each column of cats dataframe?**


## Dataframes & Vectors -- common data structures in ecology

*note: everything in a vector must be the same data type*

What if we want to check this with R?
```{r}

# use str() or class()  function 


# Option 1: check the whole dataframe


# Option 2: check specific column




# R thinks string liking is a number, but we want it to be classified as logical


# use as.logical() function


# Note: what's the difference between assigning a vector vs. typing command?



```


What if we want to add another vector to our dataframe?
```{r}
# for this example, we will need to make up a new vector, fluffy factor (1 - 5: least to most fluffy)



# Exercise 1: How many individual observations of fluffy factor  should we have?

# HINT: NEW_VECTOR <- c(OBSERVATION1, OBSERVATION2, OBSERVATION3)



str(cats)

# Exercise 2: What if we added "" around our observations?



str(cats)


# Exercise 3: let's add fluff to cats using cbind()

# HINT: NEW_DATAFRAME <- cbind(DATASET1, DATASET2)


```

**Check in: What if we added 4 observations of fluff to cats?**


How to rename vectors in dataframe
```{r}

# Exercise 1: check with names()



# Exercise 2: change only 1 name

  # let's change 'likes_string' to 'string_philic'

# HINT: names(DATASET)[COLUMN_NUMBER] <- "NEW_COLUMNNAME"



# Make sure to check the type of data after you alter it


```

How to reorder vectors that are factors
```{r}
# Exercise 1: How to reorder cat coats to be black, calico, tabby?


# HINT: DATASET$FACTOR <- factor(DATASET$FACTOR,
                                    #levels = c("LEVEL1",
                                            #   "LEVEL 2",
                                            #   "LEVEL 3"))


```


How to call vectors in dataframe
```{r}

# syntax matters

# note: [ ] vs ( )

# note: (rows, columns)

# note: , vs :

# note: | vs &


# Exercise 1: call first item 



# Exercise 2: call the 2 and 3rd item (sequential)



# Exercise 3: select all except the first using -



# Exercise 4: select 1,3 item (nonsequential)

# HINT: dataframe[c(COLUMN1,COLUMN3)]


```


How many subtle ways are there to call items in a dataframe?
```{r}
cats[1] # returns first slice of list 

cats[[1]] # returns contents of list item

cats$coat # calls column by name

cats[1,1] # calls 1st row, 1st column

cats[,1] # what do you think?

cats[1,] # what do you think?
```


## Let's try these techiques with more realistic data (aka gapminder)

#### Questions 

We will use these **functions** within the package `base`
a. dim()
b. str()
c. colnames()
d. head(), tail()


**a. How many rows and columns in the gapminder dataframe? Hint: use dim()**
```{r}
# let's look at the help page


# let's run the command


# remember: output is in the format of rows, columns

```


**b. What are the different types of data in gapminder dataframe? Hint: use str()**
```{r}

# run the command 

# HINT: FUNCTION(ARGUMENT)


# how do the outputs of str() and dim() differ?

```


**c. What are the different column names in gapminder dataframe? Hint: names() or colnames()**
```{r}

# run the command


# why do you think it's important to know the column names?


# are there any names you would like to rename?



# I personally like separating column names by _, so let's rename columns that would benefit by that


# HINT: names(DATASET)[COLUMN_NUMBER] <- "NEW_COLUMNNAME"

```



**d. What countries have the highest and lowest gdpPercap? Hint: rearrange, then select using head() or tail()**
```{r}


# Challenge

# we will need tidyverse for this 

# arrange dataframe by gdpPercap

?arrange()

ordered_gapminder <- arrange(gapminder, desc(gapminder$gdpPercap))



# run the commands using functions head() and tail()



# Challenge: How to check the last 10 rows of a dataset?

# HINT: FUNCTION(ARGUMENT 1, n = NUMBER OF ROWS)



```



## Debrief

+ If you are given a dataset, you should know **how to start to looking** at data
  + e.g. dim(), str()
  
+ With the function str(), you are able to tell **what type of data** you have

+ With the function colnames(), you know exactly what your columns/groupings are called
  + this will help with **next weeks lesson on how to manipulate data** for your own questions!




