---
title: "DataCarpentry_Day4_ControlFlow"
author: "sbsambado"
date: "9/22/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# necessary dataframes

  # Dataframe 1 -- from Carpentry
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", stringsAsFactors = TRUE) 
```
Spacing is everything - find a format that works for you!

Notes:

+ need a boolean statement
+ condition in ( )
+ make sure output for every condition

# if
if (condition is true) {
  perform action
}


Or if you have multiple conditions

# if ... else
if (condition is true) {
  perform action
} else {  # that is, if the condition is false,
  perform alternative action
}


### If statements
```{r}
x <- 8

if (x >= 10) { # has to evaluate to TRUE/FALSE
  print("x is greater than or equal to 10") 
  } else{ # for TRUE condition
  print("x is less than 10")
}

x


# want to throw in a 3 rd condition, make sure to add condition () after else if statements

if (x >= 10) { # has to evaluate to TRUE/FALSE
  print("x is greater than or equal to 10") 
  } else if (x > 5){ # for TRUE condition
  print("x is less than 5")
  }


# situationally dependent, order matters

  
  
x <- 4 == 3 # doubel equals means ASK A QUESTION
x <- FALSE
x <- TRUE

if(x) {
  "4 equals 3" # if x is TRUE
} else {
  "4 does not equal 3" # if x is FALSE, we'll get this statement
}


```

Challenge 1: Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

```{r}
dim(gapminder[(gapminder$year == 2002),]) # index it

# ^ that prints years, but we want a statement

rows2002_number <- nrow(gapminder[(gapminder$year == 2002),])


rows2002_number >= 1 # make into a boolean statement


# slow way -- evaluating all of the rows

if(nrow(gapminder[(gapminder$year == 2002),]) >= 1) {
  print("redcords for year 2002 found")
}


# faster way -- evaluating all of the rows

if(any(gapminder$year == 2002)) { # any() asks a TRUE/FALSE
  print("redcords for year 2002 found")
}
  
# what tatum would have done  

if(gapminder$year == 2002){
    print("redcords for year 2002 found")
} # can't simply give a vector of booleans, need a single boolean



ifelse() # R special

# ifelse(condition, true action, false action) 

y <- -3
ifelse(y < 0, "y is neg", "y is pos or 0")

# you can than start nesting these ifelse conditions!


all(gapminder$year == 2002) # not every value is TRUE -> gives false
any(gapminder$year == 2002) # at least one value is TRUE -> gives TRUE
  
```


### Repeating observations - eliminate the copy and paste!

for (iterator in set of values) {
  do a thing
  then, update the iterator variable & start over
}

```{r}

# basic
for(i in 1:10) { # has to be a set, can't be a single number
  print(i) # print value of i
}


# slightly less basic, nested for loops
# i, j, k for nested alphabet

for(i in 1:5) { # set i to 1 
  for(j in c("a", "b","c", 'd', 'e')) { # set j to a, j updates first
    print(paste(i,j))
  }
}

# we think it's going to be 1a but inner most loop goes first
# inner loops go thorugh all iterations
# THEN update the outer loop

# good test: set your iterator variable to one number or value, and run line by line

output_vector <- c() # create empty vector that's waiting to be filled


for(i in 1:5) {
for(j in c("a", "b","c", 'd', 'e')) { 
  temp_output <- paste(i,j) 
  output_vector <- c(output_vector, temp_output) # growing your result for smaller observations < 1000
}
}

# temp output will have the last value 


#### better option: create slots to fill

# make cubbies first and then fill with shoes

output_matrix <- matrix(nrow = 5, ncol = 5)

# speed it up by making j vector
j_vector <- c("a", "b","c", 'd', 'e')

for(i in 1:5) {
for(j in 1:5) { 
  temp_j_value <- j_vector[j] # j takes on each value of j_vector
  temp_output <- paste(i,temp_j_value)  
  output_matrix[i,j] <- temp_output
}
}

output_matrix # matrix could be helpful if you're trying to fill out a whole dataframe

# but we want vector

output_vector2 <- as.vector(output_matrix)

```

general work flow

1. set up cubbyholes (make a vector or matrix of right size, filled with NA)

2. then, in the for loop, fill each cubbyhole with a value based on the index 


### Do While loops

while(this condition is TRUE) {
do the thing
}

```{r}
z <- 1
while(z > 0.1) { # going to keep doing this forever until condiition is met
  z <- runif(1) # draw 1 random number from standard uniform
  cat(z, "/n") # mix of paste and print, /n is new line
}


```

Challenge 2: Compare the objects output_vector and output_vector2. Are they the same? If not, why not? How would you change the last block of code to make output_vector2 the same as output_vector?

```{r}

```


Challenge 3: Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller or larger than 50 years.
```{r}

## one way to solve this


for(icontininent in unique(gapminder$continent)) { # select unique values of continent
  tmp <- gapminder[gapminder$continent == icontininent,] # create temporary variable storing subset
  cat(icontininent, mean(tmp$lifeExp, na.rm = TRUE), "\n") # return calculated life exp
  rm(tmp)
}


threshold <- 50

for(icontininent in unique(gapminder$continent)) {
  tmp <- mean(gapminder[gapminder$continent == icontininent, "lifeExp"]) # make sure you specifically say lifeExp
  
  if(tmp < threshold){ # check if lifeExp is < 50
    cat("avg life exp in ", icontininent, "less than", threshold,"\n") # if < 50, print this
  } else {
    cat("avg life exp in", icontininent, "greater than", threshold, "\n")
  } # end if else condition
  rm(tmp) # remove temporary to avoid weird memory storage stuff
}





```

