---
title: "DataCarpentry_Day2_ExploringDataframes"
author: "Imani Russell"
date: "9/21/2021"
output: html_document
---

# Adding columns and rows in data frames

REMEMBER: data frame columns = vectors
+ add a new column by creating a vector

## Columns
```{r}
# create a new vector
age <- c(2, 3, 5)

# reload cats dataframe
cats <- data.frame(coat = c("calico", "black", "tabby"), 
                    weight = c(2.1, 5.0, 3.2), 
                    likes_string = c(1, 0, 1)) 

cats

# add new column to cats
cbind (cats, age)

# REVIEW: what happens if we try to add this vector as a new column?
age <- c(2, 3, 5, 2)
cbind (cats, age)

# check number of rows in cats
nrow(cats)
dim(cats)

# check length of age (nrow vs length)
length(age)

# overwrite old cats dataframe with new one including age
age <- c(2, 3, 5) # reassign age

cats <- cbind (cats, age)
```


## Adding rows

Each row = a list (*lists have elements of different types vs vectors = same data type)
```{r}
# create a new list
newRow <- list ("tortoiseshell", 3.3, TRUE, 9)

# add new row to dataframe
cats <- rbind (cats, newRow) # warnings vs errors

cats
```

Why do we get NA instead of tortoiseshell?

# Factors

Factor = data type in which each value represents a level

Here, coat = factor with 3 levels: "black", "calico", "tabby".

+ R will only accept values that match one of these levels
+ Solution: add "tortoiseshell" as a possible factor level

```{r}
# check how many levels are in the factor coat 
levels(cats$coat)

# add "tortoiseshell"
levels(cats$coat) <- c(levels(cats$coat), "tortoiseshell")

# add new row to dataframe
cats <- rbind (cats, list ("tortoiseshell", 3.3, TRUE, 9))
```

ALTERNATIVE: change cats$coat from factor to character

+ PRO: can add any word to the column
+ CON: lose handy categories/functionality of factors
```{r}
# check to see class of coat variable
str(cats) 
class(cats$coat) # if you don't want to see the class of everything or detail
#typeof(cats$coat)

# reassign as character
cats$coat <- as.character(cats$coat)

str(cats)
```


CHALLENGE 1:
1 cat year = 7 human years

1) create a vector called human_age by multiplying cat$age by 7
2) convert human_age to a factor
3) convert human_age back to a numeric vector using as.numeric(). Divide that by 7 to get the original ages back. Explain what happened.



```{r}
# 1
human_age <- cats$age * 7
str(human_age)
class(human_age)

# 2
human_age <- factor(human_age)
human_age <- as.factor(human_age) 
human_age <- as.character(human_age)

str(human_age)

# 3
human_age <- as.numeric(human_age)
str(human_age)

# look back and compare original human_age with current
# 1 2 3 4 4 4 were originally the factors, which are stored as integers, with labels = the values (here, 28, 35, 56, and 63)
# changing factors to numeric returns the underlying integers and not the values/labels we're interested in
# to return the original numbers, we should convert human_age to class character (as.character()) and then to numeric.

human_age / 7

# when using read.csv(), remember to set your argument stringsAsFactors = FALSE to prevent typos from affecting your factor conversions
```



# Removing Rows

Look back at cats dataframe - we now have an extra row we don't need. Let's remove it!
```{r}
cats

cats [-4, ] # [row, column] remove the 4th row but leave columns alone

cats [ c(-4, -5), ] # remove rows 4 and 5

na.omit (cats) # remove all rows with NAs - be careful using this function!!!

# reassign dataframe so we save our changes
cats <- na.omit (cats)
```


# Removing columns

```{r}
# remove age column
cats [, -4] # keep all rows but remove column 4

# alternative
drop <- names (cats) %in% c("age") # do any data frame column names appear in "age"? or whatever word you enter

cats [, !drop] # return all rows and all columns that are not TRUE in drop (remember TRUE = 1, FALSE = 0)
```


# Appending to a data frame

Can also bind 2 data frames together using rbind()
```{r}
cats <- rbind (cats, cats)
cats

rownames(cats)

# rownames get a little crazy - remove them and let R rename them sequentially
rownames(cats) <- NULL

cats
```

CHALLENGE 2: 

You can use the  following syntax to create a dataframe:

df <- data.frame(id = c("a", "b", "c"),
                 x = 1:3,
                 y = c(TRUE, TRUE, FALSE),
                 stringsAsFactors = FALSE)
                 
Make a data frame that holds the following information:

+ first name
+ last name
+ lucky number

Use rbind() to add an entry for someone else (family member, someone else in the workshop, etc!).

Use cbind() to add a column with each person's answer to the question: "Is it time for a coffee break?"

```{r}
df <- data.frame(first = c("Grace"),
                 last = c("Hopper"),
                 lucky_number = c(0),
                 stringsAsFactors = FALSE)
df <- rbind(df, list("Marie", "Curie", 238) )
df <- cbind(df, coffeetime = c(TRUE,TRUE))
```


# Realistic example

Read in the gapminder dataset
```{r}
gapminder <- read.csv ("data/gapminder_data.csv", stringsAsFactors = FALSE)

# if you weren't here yesterday!
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", stringsAsFactors = TRUE) 
```

let's explore the data a little bit!
```{r}
str(gapminder) # review

summary(gapminder) # returns a few summary statistics/descriptive summary

# factors: by number of items in each level
# numeric: quartiles, mean, etc
# character: length, class, mode

# you can check summary for vectors too
summary (gapminder$country)


# look at individual columns
typeof(gapminder$year) # what does R see this as?
class(gapminder$year) # what do functions read this as?

typeof(gapminder$country)

str(gapminder$country) 

# double check dimensions, etc
length(gapminder) # gives number of columns, not rows!

typeof(gapminder) # remember, dataframe = list of vectors and factors

ncol(gapminder) # does the same thing - number of columns

colnames(gapminder) # what are these column names?

nrow(gapminder) # number of rows

dim(gapminder) # both! remember: rows, columns
```

CHALLENGE 3:

It’s good practice to also check the last few lines of your data and some in the middle. How would you do this?

Searching for ones specifically in the middle isn’t too hard, but we could ask for a few lines at random. How would you code this?

```{r}
# check last 6 (default)
tail(gapminder)


# check last 15
tail(gapminder, n = 15)



# get a few random lines
gapminder[sample(nrow(gapminder), 5), ] # nested function:P function passed as an argument to another function

# within gapminder, we want the full rows that meet the following conditions:
# we want those rows to be a random sample, taken across all possible rows of gapminder

# may help to break down into parts:
# gapminder[x, ]
# gapminder[sample (x), ]
# gapminder[sample (nrow (x), 5), ]
```

**There are often multiple ways to do the same thing in R!**


CHALLENGE 4

Go to file -> new file -> R script, and write an R script to load in the gapminder dataset. Put it in the scripts/ directory and add it to version control.

Run the script using the source function, using the file path as its argument (or by pressing the “source” button in RStudio).

```{r}
source(file = "scripts/load-gapminder.R")
```




CHALLLENGE 5

Read the output of str(gapminder) again; this time, use what you’ve learned about factors, lists and vectors, as well as the output of functions like colnames and dim to explain what everything that str prints out for gapminder means. If there are any parts you can’t interpret, discuss with your neighbors!








