---
title: "DataCarpentry_Day3_FunctionsExplained"
author: "tatum katz"
date: "9/22/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---
## Functions Explain -- Software Carpentry Lesson

Notes

+ functions give us a name we can use again
+ define set of input, know what output looks like


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Dataframe 2 -- gapminder (from Carpentries)

gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", stringsAsFactors = TRUE) 


```


### Basics of functions
```{r}
fahr_to_kelvin <- function(temp) { # name the function, identify the arguments
  kelvin <- ((temp - 32)) * (5/9) + 273.15 # actual function
  return(kelvin) # return a result
}


fahr_to_kelvin(32)

## Challenge 1
# Write a function called kelvin_to_celsius() that takes a temperature in Kelvin and returns that temperature in Celsius.

kelvin_celsius <- function(temp){
  celsius <- ((temp - 273.15))
  return(celsius)
}

kelvin_celsius(284)


## Challenge 2
# Define the function to convert directly from Fahrenheit to Celsius, by reusing the two functions above (or using your own functions if you prefer).


fahr_celsius <- function(temp){
  temp_kelvin <- fahr_to_kelvin(temp)
  result <- kelvin_celsius(temp_kelvin)
  return(result)
}


fahr_celsius(120)

```



### Defensive programming

+ prevent issues from occuring
```{r}
# let's make fahr_to_kelvin more defensible

fahr_to_kelvin <- function(temp) { # name the function, identify the arguments
  if(!is.numeric(temp)){ # temp must be numeric
    stop("temp must be a numeric vector")
  }
  kelvin <- ((temp - 32)) * (5/9) + 273.15 # actual function
  return(kelvin) # return a result
}

# test it

fahr_to_kelvin('32') # wrong
fahr_to_kelvin(32) # right


# second modification -- stop if not

fahr_to_kelvin <- function(temp) { # name the function, identify the arguments
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32)) * (5/9) + 273.15 # actual function
  return(kelvin) # return a result
}


##  Challenge 3
# Use defensive programming to ensure that our fahr_to_celsius() function throws an error immediately if the argument temp is specified inappropriately.

fahr_celsius <- function(temp){
  stopifnot(is.numeric(temp))
  temp_kelvin <- fahr_to_kelvin(temp)
  result <- kelvin_celsius(temp_kelvin)
  return(result)
}

fahr_celsius('g') # test 
```

### Combining functions
```{r}

calcGDP <- function(dat) {
  gdp <- dat$pop * dat$gdpPercap
  return(gdp)
}

calcGDP(head(gapminder)) # this is useless, let's build it up



# build up 

calcGDP <- function(dat, year = NULL, country = NULL) { # create the function, null are default
  # arguments:
  # dat: required! a dataset with pop and gdpPercap
  # year: optional, a numeric vector, a year
  # country: opetion, character vector, a country
  if(!is.null(year)) { # if year is specified
    dat <- dat[dat$year %in% year, ] # subset out just the user-specified year
  }
  if(!is.null(country)) { # if country is specified
    dat <- dat[dat$country %in% country,] # and, if so, subset out just that country 
  }
  
  gdp <- dat$pop * dat$gdpPercap # then, calculate the gdp for each row in the subset
  new <- cbind(dat, gdp = gdp) # stick that gdp on as a new column in the dataset
  return(new) # return your enw dataframe with the gdp
}

calcGDP(dat = gapminder, year = 2002) # all countries in 2002

calcGDP(dat = gapminder, year = 2002, country = 'Afghanistan')
```

### How to use source
```{r}
source("functions/functions-lessons.R") # all your functions can be uploaded

calcGDP(gapminder, year = 1997)

```


More notes:

+ vectors in functions only exist in functions, not global environment
+ subset data within function are considered passby values


Challenge 4: Test out your GDP function by calculating the GDP for New Zealand in 1987. How does this differ from New Zealand’s GDP in 1952?

```{r}
# long version
calcGDP(gapminder, country = "New Zealand", year = 1987)
calcGDP(gapminder, country = "New Zealand", year = 1952)

# short version

calcGDP(gapminder, year = c(1952, 1987), country = "New Zealand")



```

Challenge 5: The paste() function can be used to combine text together, e.g:
```{r}
# paste function is used to combine text together

best_practice <- c("Write", "programs", "for", "people", "not", "computers")
paste(best_practice, collapse=" ")

# Write a function called fence() that takes two vectors as arguments, called text and wrapper, and prints out the text wrapped with the wrapper:

fence <- function(text, wrapper) {
  text <- c(wrapper, text, wrapper)
  output <- paste0(text, collapse = " ") # paste0 pastes together with no spaces, collapse gives one long string
  return(output)
}

fence(text=best_practice, wrapper="***")


fence(text = "this is so freakin fun", wrapper = "!!!!")


```

When you first start out, your workflow will probably look a lot like this:

1. Write a function
2. Comment parts of the function to document its behaviour
3. Load in the source file
4. Experiment with it in the console to make sure it behaves as you expect
5. Make any necessary bug fixes
6. Rinse and repeat.

